---
title: "Chapter 7: transform & visualise"
author: "Elise Gould"
date: "03/02/2017"
output: github_document
---

```{r include=FALSE, eval=TRUE}
library(tidyverse)
bat_dat <- read_csv("../data/bat_dat.csv")
richness_by_season <- bat_dat %>% 
        select(Site, Habitat, Season, BatSpecies) %>%
        group_by(Habitat, Season) %>% #For each habitat
        summarise(mean_richness = mean(BatSpecies),
                  sd_richness = sd(BatSpecies),
                  n_richness = n()) %>%
        mutate(se_richness = sd_richness / sqrt(n_richness))
```


# Ggplot2

Ggplot is based on a 'grammar of graphics', the idea that you can build every graph from the same few components:

- a data set
- a set of **geoms** (visual markers representing data points)
- And a coordinate system, a method of arranging points in space

To display data values, variables in the dataset are mapped to aesthetic properties of the specified **geom**, such as size, colour and x, y locations. This creates an empty plot. Let's try with the iris data:

```{r}
library(ggplot2)
ggplot(data = iris, aes(x = Species, y = Petal.Width))
```

We add layers to the plot using the `+`sign:
```{r}
ggplot(data = iris, aes(x = Species, y = Petal.Width)) +
        geom_point(aes(colour = Species)) +
        theme_bw()
```

# Challenge, create a ggplot:






# Final Challenge (easy): Create a Plot with ggplot2

Walk through the following steps to create a ggplot in your Rmarkdown

1. create a new chunk, call it `plot`, and give it the option `echo = FALSE`
2.  Using the richness_by_season data, establish an empty plot providing the following aesthetics:

- x: Habitat
- y: mean_richness
- fill: Season
3. Add the following layers with the plus sign to create a bar plot with error bars:
```
geom_bar(stat = "identity", position = position_dodge())

geom_errorbar(aes(ymin = mean_richness - se_richness, 
                  ymax = mean_richness + se_richness), 
                  width = 0.1,
                  position = position_dodge(0.9))
```

```{r plot-answer, eval = FALSE, include=FALSE}
ggplot(data = richness_by_season, 
       aes(x = Habitat, y = mean_richness, fill = Season)) +
        geom_bar(stat = "identity", position = position_dodge()) +
        geom_errorbar(aes(ymin = mean_richness - se_richness, 
                          ymax = mean_richness + se_richness), 
                      width = 0.1,
                      position = position_dodge(0.9)) # shifts the error bars over the correct bar
```

# Resources:

[Ggplot2 cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/12/ggplot2-cheatsheet-2.0.pdf)
